Data-driven approaches to identifying knowledge gaps in marine invertebrates biodiversity

### 2017-03-08 manuscript.Rmd
edits to manuscript

---
#author: Fran\c{c}ois Michonneau
title: A data-driven approach to identifying knowledge gaps in marine invertebrates biodiversity
output: pdf_document
---

# Summary

Species identification from DNA sequences is poised to revolutionize our
understanding of biodiversity, to facilitate environmental monitoring, and to
disentangle trophic networks. However, this approach relies on comprehensive
databases of accurately identified sequences.  Since its introduction in the
early 2000's, DNA barcoding, the sequencing of a common locus across taxa
bearing species-specific information, has revitalized taxonomy by facilitating
species delineation. More recently, advances in high-throughput sequencing
technologies have enabled the parallel sequencing of thousands of species that
occur in environmental samples. The approach initially developed to study
microbial communities such as the ones from guts or soil, has been modified and
extended to be applied to metazoan communities. This approach opens new research
venues as it becomes possible to estimate the diversity from species-rich
samples that were virtually inaccessible before. This is particularly true in
marine communities that include multiple phyla and organisms varying in size by
order of magnitude, making identification challenging. However, the utility of
the data generated by these projects is enhanced by the possibility of matching
the DNA sequences obtained to previously sequenced and correctly identified
species. Here, we combined information available in public biodiversity
databases to assess the proportion of identified species that have been
sequenced in the continental United States. We show that the effort is
consistent across the better-studied taxa, with `r average_prop_species_seq`% of
the fauna sequenced. Using a DNA barcode dataset of `r n_larvae` planktonic
larvae, representing `r n_esus` species, collected in coastal waters of Florida,
we find a similar proportion of species that can be assigned to a previously
sequenced species ( `r format_output(100 * prop_plankton_with_match)`%).  Our
results show that most marine species remain to be sequenced. Achieving this
goal will required a concerted effort, and resources.

<!---
the content of insect traps, marine sediments, plankton
samples, or communities settling on artificial substratum. When applied to
metazoan communities, these sequencing approaches typically target the barcoding
gene and theefore has been dubbed "meta-barcoding". These applications are also
making the public databases for these DNA barcodes all the more valuable. To
deliver on the meta-barcoding promises, it is crucial to have comprehensive
taxonomic coverage represented in the DNA sequences public databases. Here, we
investigate the proportion of species for each marine phylum that have been
sequenced at various spatial scale.
--->

@@

With increased anthropogenic pressure on ecosystems, it is crucial to become
efficient at documenting biodiversity before species go extinct. This is
especially important in marine ecosystems where an undefined, but large,
fraction of the biodiversity remains to be discovered especially among
invertebrates. However, taxonomic expertise is dwindling, and there are now only
a handful of natural history museums with active marine invertebrates
collections. Museum collections have been recording our knowledge about species
and their geographical ranges, but until recently, this information was
fragmented across institutions and difficult to synthesize. With collection
digitization and other large-scale projects have aimed at gathering biodiversity
information, public databases can now be used to enable synthesis and highlight
knowledge gaps. Here we use these data to synthetic the state of knowledge for
marine invertebrates, and demonstrate that the stage is set to combine these
data sources to inform where future efforts for documenting biodiversity.

We combined `r nrow(global_idigbio_species_list)` museum records for marine
invertebrates collected in the continental US representing
`r length(store_idigbio_records()$list())` phyla and
`r length(unique(global_idigbio_species_with_worms_info$worms_valid_name))`
species. Most samples were collected on the continental
shelf (Fig. 1). For the East coast, species richness is the highest in South Florida but
it is relatively uniform along the coastline. For the West coast, the Salish
Sea, the Bay Area, and Southern California have the highest richness (Fig. 2).

The number of sampled species is vastly underestimated.
`r prop_singleton_species`% of the species are known from a single
record. Futhermore, the strong correlation between the number of samples
collected and the species richness suggests that any new sampling effort
will lead to new records (Fig. 3).

DNA barcoding is one approach that can be used to alleviate the discrepancy
between the increased need for biodiversity inventories while taxonomic
expertise is dwindling. DNA barcoding aims at assembling a comprehensive
database of DNA sequences that are associated with accurately identified
vouchered specimens. Future identifications can then rely on this database. DNA
barcoding can also facilitate species delineation, in turn, increasing the
accuracy of the identifications.

Recent developments in high-throughput sequencing technologies are giving new
applications to these DNA barcode databases. Meta-barcoding enables the parallel
sequencing of thousands of species that occur in environmental samples.  This
approach opens new research venues as it becomes possible to estimate the
composition of species-rich samples that were virtually inaccessible before
(e.g., fish gut content, marine sediment samples, plankton tows). The ability to
use DNA to identify members of these communities relies on comprehensive DNA
barcode databases. To assess the proportion of recorded species in the US that
has been sequenced, we searched for each species occurring in museum collections
for possible matches in BOLD (which also includes GenBank records). We found
that the proportion of sequenced species is consistent across the better-studied
taxa (overall `r average_prop_species_seq`%, Fig. 4). This proportion is consistent
across phyla despite important differences in species richness, and taxonomic
tradition. We also found that phyla representative of the meiofauna are
under-represented, and very few specimens and sequences are available.

Identified specimens from museum collections could be used to complete these
datasets. Traditional approaches relying on Sanger sequencing might be limiting as
many species have not been collected recently enough to guarantee good quality DNA
(`r prop_species_not_collected_since_1990`% have not been collected since
1990). However, high-throughput sequencing could offer alternatives to recover
sequences from older specimens.

Besides data-mining museum records to identify knowledge gaps, new sampling
approaches can be used in the field to rapidly assess how much of the fauna is
known. Meta-barcoding and environmental DNA approaches appear promising to
sample _en masse_ species that are found in a given location. Because most
marine invertebrates have a planktonic larval stage, they may co-occur in the
water column, allowing to sample species across habitats. To illustrate this
approach, we individually DNA barcoded `r n_larvae` planktonic larvae for DNA
barcoding, collected over the course of year, representing `r n_esus` species,
collected in coastal waters of Florida. We found that
(`r format_output(100 * prop_plankton_with_match)`%, Fig. 4) can be assigned
to a previously sequenced species. Similar proportions have been obtained from
individually barcoded species that occur on ARMS. However, this proportion
dropped to less than 5% for smaller organisms sequenced with meta-barcoding.

The species identifications in BOLD are far from perfect. Of the
`r sum(prop_plankton_with_species_match_by_phylum$n_esu_match)` species that have a
match in public database,
`r format_output(prop_plankton_with_multi_species_match * 100)`% have been
identified as more than one taxon in BOLD. A few of these records might represent
species complexes, but most illustrate the difficulty to identify precisely
marine invertebrates.

<!-- Studying biodiversity and identifying knowledge gaps about species and their -->
<!-- distributions need to become more data-driven to target under-studided areas and -->
<!-- taxonomic groups. To survive the biodiversity crisis, we need -->
<!-- inter-disciplinarity among -->
<!-- scientists  only need to be taxonomists that can reliably -->
<!-- identify species, but also molecular biologists. -->

![Number of museum records for marine inverterbrates](figures/map_sampling_effort.pdf)

@@

![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.png)

![Proportion of species occuring in natural history collections that have been DNA barcoded](figures/proportion_barcoded.pdf)

@@

>>>>

---
#author: Fran\c{c}ois Michonneau
title: "Data-driven approaches to identifying knowledge gaps in marine invertebrates biodiversity"
output:
  pdf_document #:
    #pandoc_args: ["-F","pandoc-crossref"]
---

<!-- use Pimm+2014 intro as model on using short informative sentences to build -->
<!-- 1st paragraph. -->

<!-- Species identification from DNA sequences is poised to revolutionize our understanding of biodiversity, to facilitate environmental monitoring, and to disentangle trophic networks. However, this approach relies on comprehensive databases of accurately identified sequences.  Since its introduction in the early 2000's, DNA barcoding, the sequencing of a common locus across taxa bearing species-specific information, has revitalized taxonomy by facilitating species delineation. More recently, advances in high-throughput sequencing technologies have enabled the parallel sequencing of thousands of species that occur in environmental samples. The approach initially developed to study microbial communities such as the ones from guts or soil, has been modified and extended to be applied to metazoan communities. This approach opens new research venues as it becomes possible to estimate the diversity from species-rich samples that were virtually inaccessible before. This is particularly true in marine communities that include multiple phyla and organisms varying in size by order of magnitude, making identification challenging. However, the utility of the data generated by these projects is enhanced by the possibility of matching the DNA sequences obtained to previously sequenced and correctly identified species. Here, we combined information available in public biodiversity databases to assess the proportion of identified species that have been sequenced in the continental United States. We show that the effort is consistent across the better-studied taxa, with ` average_prop_species_seq`% of the fauna sequenced. Using a DNA barcode dataset of ` n_larvae` planktonic larvae, representing ` n_esus` species, collected in coastal waters of Florida, we find a similar proportion of species that can be assigned to a previously sequenced species ( ` format_output(100 * prop_plankton_with_match)`%).  Our results show that most marine species remain to be sequenced. Achieving this goal will required a concerted effort, and resources. -->

<!---
the content of insect traps, marine sediments, plankton
samples, or communities settling on artificial substratum. When applied to
metazoan communities, these sequencing approaches typically target the barcoding
gene and therefore has been dubbed "meta-barcoding". These applications are also
making the public databases for these DNA barcodes all the more valuable. To
deliver on the meta-barcoding promises, it is crucial to have comprehensive
taxonomic coverage represented in the DNA sequences public databases. Here, we
investigate the proportion of species for each marine phylum that have been
sequenced at various spatial scale.
--->

<!-- studies that have summary statistics on what has been sequenced for various marine phyla:
- for Antarctica: Grant+Linse2009-Polar Biology:

We know that some perturbations have large effects because they are visible, but what about perturbations for which we don't see the effects directly because they affect small species? :: checkout nature paper on sea grasses.

- Describing new species is crucial but not sufficient. Without knowing their ranges, and their habitats, we can't study or predict how perturbations of the environment are going to affect their distributions and their susceptibility to perturbations. The process of figuring out the ranges of species is more time consuming than possibly finding new species; and it requires shared knowledge in the community about these new species. Using barcoding and other moleuclar techniques to improve the documentation of species distributions is going to be critical.

There can be public involvement in species description, but the exploration and collecting may remain an affair of specialists as it requires manipulating technical gear of have expertise. There are still many species that are easy to access to discover, but the bigger gap is in the species that are difficult to access, and small.

We also need to better coordination in describing new species; so that efforts are not duplicated.

Possibly also need to set guidelines on how to treat cases where the molecular data alone is too ambiguous.

-->

@@

<!--  study focused on USA but conclusions more general given that it's arguably one of the better studied area + data availability -->
<!-- * barcoding has shown that typically marine invertebrates are complex of cryptic species with small ranges; therefore potentially more vulnerable to extinction risk. -->
<!-- * need collaboration among researchers to share data standards, protocols -->
<!-- * use existing knowledge to fill missing gaps -->
<!-- * running against time -->
<!-- * we need to document species diversity better for conservation (knowing what we protect and to measure the effectiveness of the protection) -->
<!-- * how do we conserve animal species that we can't see, we don't find pretty, don't have eyes -->
<!-- * recent advances in technologies make possibility of developing a global monitoring program possible: reduced costs of sequencing, detection of rare species, etc.... -->
<!-- * Using this kind of approaches at a large scale might also be a good way to have a better sense of how much of the biodiversity we are missing. -->
<!-- * Most species are rare, especially in very diverse places,  and the degradation of habitats makes it even more difficult to find these rare species. -->

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges, detect changes in the composition of communities, and more generally accelerate the documentation of species diversity. This is especially pressing in marine ecosystems where a significant portion of the biodiversity remains to be discovered and described.

Until recently our knowledge of species and their distributions was difficult to synthesize because it was scattered and difficult to access. However, recent advances in digitization and public databases efforts have made widely available the content of natural history collections, publications about species, their geographical distributions, and their genetic data. Given the urgency with which we need to document biodiversity and the limited means available to do so, we need to use and build on this existing knowledge to identify gaps, and to be more efficient at collecting information and understanding our natural world.

The incorporation of molecular data in the study of biodiversity has highlighted the pervasiveness of cryptic species, and generally the fact that species have smaller ranges than their taxonomic identity implied. Notably, the wide adoption of DNA barcoding for the study of marine diversity has clarified our understanding of species limits in many groups, but has also enabled the identification of species during their larval stages (when the lack of morphological features make species identification difficult). DNA barcoding has also been seen as a possible solution to the widening gap between the increased need for biodiversity inventories while taxonomic expertise is dwindling. By matching DNA barcodes of unidentified species to a public database of accurately identified and vouchered specimens, one can greatly speed up the process of species identification by matching them to already identified species. Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.

Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases. Meta-barcoding enables the parallel sequencing of thousands of DNA barcodes for species that occur in environmental samples. This approach opens new research venues as it becomes possible to estimate the composition of species-rich samples that were virtually inaccessible before (e.g., gut content, marine sediment samples, plankton tows). For instance, by sequencing the plankton of coastal areas, it becomes possible to sample the larval stages of many species that would otherwise be difficult to sample. However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.

With the cost of sequencing going down, and the prospect of having an automated sampling process, these technologies will soon be used for ecosystem and biodiversity monitoring. However, to interpret the results and understand how species distributions shift with global change or other perturbations, we cannot only rely on the identity of the species, but we also need to take advantage of existing knowledge about these species that is coming from natural history collections. In any case, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included public databases.

<!-- here there are 3 things: data availability (having good APIs to access the -->
<!-- data); data comprehensiveness (how many species are captured in the -->
<!-- database); data accuracy (can we trust the data that is found in the -->
<!-- databases) -->

Here, we assess the level of completeness of biodiversity public databases for marine invertebrates that occur within the boundaries of the EEZ of the US. Specifically, we use comprehensive regional faunistic lists to assess how much of this knowledge is captured by current digitization efforts of museum collections and occurrence records; and we estimate how many of the species have genetic data publicly available that would be useful in the context of identifying samples from meta-barcoding studies.

We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea. Species are unevenly represented in databases, with 25% to 50% of species listed to occur in these locations, that are not represented in digital records of museum collections. Typically a higher proportion of species from the Pacific Coast than the Gulf of Mexico are not represented. Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species known to occur in the US have digitized museum specimens.

The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla. About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.  Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographical coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought. Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species. It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to ensure that the extent of their genetic diversity is captured. This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN. This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.

<!-- However, among these records, xx% have been sampled in the US (xx%), Mexico (xx%) or Canada (xx%).  -->
<!-- Paragraph about the same numbers but based on species list from iDigBio: -->

<!-- - how many species are in iDigBio and don't show up in species lists from Kozloff and GOM -->
<!-- - how many species in iDigBio have barcodes -->
<!-- - how many species in iDigBio have OBIS and GBIF records -->

We combined `r nrow(idigbio_records)` museum records for marine invertebrates collected in the continental US representing `r length(unique(tolower(idigbio_records[["data.dwc:phylum"]])))` phyla and `r length(unique(idigbio_records$worms_valid_name))` species. Most samples were collected on the continental shelf (Fig. 1). For the East coast, species richness is the highest in South Florida but it is relatively uniform along the coastline. For the West coast, the Salish Sea, the Bay Area, and Southern California have the highest richness (Fig. 2).

These patterns highlight our lack of knowledge about the marine fauna: `r prop_singleton_species`% of the species are known from a single record. Furthermore, the strong correlation between the number of samples collected and the species richness suggests that any new sampling effort will lead to new records (Fig. 3). This is also reflected in the cumulative numbers of species recorded for the first time and the number of samples through time (Fig. 4). The slowing down of collections in the 1970 has been accompanied with a slowing down in the discovery of new records: `r n_spp_comparison$n_spp_2017_2000` species have been documented for the first time since 2000, while `r n_spp_comparison$n_spp_1970_1960` have been documented between 1960 and 1970.

There are important discrepancies in our level of knowledge across phyla. While larger, and accessible groups are well represented in museum collections, and have a reasonable proportion of their known species represented in DNA barcode databases, the small organisms of the meiofauna are virtually absent: 12 of the 25 phyla represented in the dataset have less than 500 specimens in museums.

However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections. Museum collections have been recording our knowledge about species and their geographical ranges, but until recently, this information was fragmented across institutions and difficult to synthesize. With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps. Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.

Based on these results, we recommend that additional field expeditions are organized to sample more thoroughly the marine environment. The data from museum collections make clear that any additional samples will contain new species and new records. However, taxonomists are needed to identify and describe these new species. The samples collected during these expeditions need to be photographed to document live appearance that would otherwise be lost during preservation. We also need to capture metadata associated with the specimens to be able to understand better the physical characteristics of the habitats for these species. The specimens need to be sub-sampled and have tissues suitable for genomic analysis preserved, so we can sequence their DNA barcode, or other genomic resources.

Using metabarcoding is a promising approach to screen an habitat for its potential for species that are not found in public database, to maximize the usefulness in the future of the data generated by an expedition. Metabarcoding approaches need to complement and take advantage of the DNA barcodes already available. Additionally, in the future, data generated by metabarcoding efforts can be re-interpreted as more DNA barcodes become available. Building the library of DNA barcodes in parallel of metabarcoding efforts also ensures that matches in the database will correspond to actual specimens deposited in natural history collections; ruling out possible artifacts of the method. The knowledge will also build on existing natural history data rather than having to restart from scratch. Beyond their utility in documenting diversity, metabarcoding is also promising for understanding the functioning of ecosystems.

<!-- Paragraph on how sampling effort has changed through time. -->

<!-- - get numbers of species described by year, look if there is a correlation between sample numbers and number of new species (possible lag with time for description) -->
<!-- - see if proportion of species identified at genus (or higher) only has increased recently -->

![Number of museum records for marine invertebrates](figures/map_sampling_effort.pdf)

@@

![Relation between the number of specimen and the number of species collected](figures/plot_sampling_effort.pdf)

![Cumulative number of samples through time](figures/cum_samples_through_time.pdf)

![Cumulative number of species recorded for the first time through time](figures/cum_spp_through_time.pdf)

![Proportion of species occurring in natural history collections that have been DNA barcoded](figures/proportion_barcoded.pdf)

![Level of digitization for the Salish Sea](figures/plot_richness_per_db_koz.pdf)

![Level of digitization for the Gulf of Mexico](figures/plot_richness_per_db_gom.pdf)

@@

Identified specimens from museum collections could be used to complete these datasets. Traditional approaches relying on Sanger sequencing might be limiting as many species have not been collected recently enough to guarantee good quality DNA (` prop_species_not_collected_since_1990`% have not been collected since 1990). However, high-throughput sequencing could offer alternatives to recover sequences from older specimens.

Besides data-mining museum records to identify knowledge gaps, new sampling approaches can be used in the field to rapidly assess how much of the fauna is known. Meta-barcoding and environmental DNA approaches appear promising to sample _en masse_ species that are found in a given location. Because most marine invertebrates have a planktonic larval stage, they may co-occur in the water column, allowing to sample species across habitats. To illustrate this approach, we individually DNA barcoded ` n_larvae` planktonic larvae for DNA barcoding, collected over the course of year, representing ` n_esus` species, collected in coastal waters of Florida. We found that (` format_output(100 * prop_plankton_with_match)`%, Fig. 4) can be assigned to a previously sequenced species. Similar proportions have been obtained from individually barcoded species that occur on ARMS. However, this proportion dropped to less than 5% for smaller organisms sequenced with meta-barcoding.

The species identifications in BOLD are far from perfect. Of the ` sum(prop_plankton_with_species_match_by_phylum$n_esu_match)` species that have a match in public database, ` format_output(prop_plankton_with_multi_species_match * 100)`% have been identified as more than one taxon in BOLD. A few of these records might represent species complexes, but most illustrate the difficulty to identify precisely marine invertebrates.
